Clustering in a hyperbolic model of complex networks.

### 2019-03-13 introduction.tex
added observation section and changed the campbell-mecke section

In addition to the clustering coefficient, we shall also be interested in the {\em clustering function}.
This assigns to each integer $k$ the value
\[ 
	c(k; G) := \begin{cases}
		\displaystyle \frac{1}{N_k} \sum_{v \in V(G), \atop \text{deg}(v)=k}  c(v),  &\mbox{ if } N_k \ge 1,\\
		0, &\mbox{else,}
	\end{cases}
\]
where $N_k$ denotes the number of vertices of degree exactly $k$ in $G$. In other words, the clustering function assigns to the integer $k$ the average of the local clustering coefficient over all vertices of degree $k$. We remark that, while it might seem natural to consider $c(k)$ to be ``undefined'' when $N_k=0$, we prefer to use the above definition for technical 
convenience.  This way $c(k; G(n;\alpha,\nu) )$ is a plain vanilla random variable and we can for instance compute its moments without any issues.

@@

For a sequence $(X_n)_n$ of random variables, we write $X_n \xrightarrow[n\to\infty]{\Pee} X$ to denote that $X_n$ converges in probability to $X$, as $n \to \infty$. %In addition, $X_n \xrightarrow[n\to\infty]{L^1} X$ denotes convergence in expectation, i.e. $\Exp{|X_n - X|} \to 0$ as $n \to \infty$, which is a stronger notion than convergence in probability.

@@

\begin{theorem}\label{thm:mainktoinfty}
Let $\alpha>\frac12, \nu>0$ be fixed and let $k_n$ be a sequence of integers
satisfying $1 \ll k_n \ll n^{1/(2\alpha+1)}$. Then, writing $G_n := G(n;\alpha,\nu)$, we have
\[
	\E[|c(k_n;G_n)-\gamma(k_n)|] = \smallO{\gamma(k_n)},
\]
as $n \to \infty$, where $\gamma(\cdot)$ is as in Theorem~\ref{thm:mainkfixed}.
\end{theorem}

In particular, Theorem~\ref{thm:mainktoinfty} implies that
\[
	\frac{c(k_n; G_n)}{\gamma(k_n)} \xrightarrow[n\to\infty]{L^1} 1,
\]
so that the clustering function of the KPKVB model scales as $\gamma(k)$ as $k$ grows.

\PvdH{ToDo: add remark on maximum scaling for $k_n$.}

@@

In particular this contradicts the scaling conjectured in~\cite{krioukov2010hyperbolic} for $\alpha \leq \frac34$, and confirms it for $\alpha > \frac34$.

@@

@@

The remaining sections are devoted to the proof of Theorem~\ref{thm:mainktoinfty}, which turns out to be a lot more involved. In particular, since we push the possible scaling of $k_n$ to its maximum, a great deal of work is needed to properly control the arising error terms.

>>>>

In addition to the clustering coefficient, we shall also be interested in the {\em clustering function}.
This assigns to each integer $k$ the value
\begin{equation}\label{eq:def_clustering_function}
	c(k; G) := \begin{cases}
		\displaystyle \frac{1}{N_k} \sum_{v \in V(G), \atop \text{deg}(v)=k}  c(v),  &\mbox{ if } N_k \ge 1,\\
		0, &\mbox{else,}
	\end{cases}
\end{equation}
where $N_k$ denotes the number of vertices of degree exactly $k$ in $G$. In other words, the clustering function assigns to the integer $k$ the average of the local clustering coefficient over all vertices of degree $k$. We remark that, while it might seem natural to consider $c(k)$ to be ``undefined'' when $N_k=0$, we prefer to use the above definition for technical 
convenience.  This way $c(k; G(n;\alpha,\nu) )$ is a plain vanilla random variable and we can for instance compute its moments without any issues.

@@

For a sequence $(X_n)_n$ of random variables, we write $X_n \xrightarrow[n\to\infty]{\Pee} X$ to denote that $X_n$ converges in probability to $X$, as $n \to \infty$.

%We further adopt standard notations on asymptotic behavior of functions and sequences. That is, for two functions $f$ and $g$, we write $f(n) = \smallO{g(n)}$ as $n \to \infty$ if $\limsup_{n \to \infty} f(n)/g(n) = 0$ and $f(n) = \bigO{g(n)}$ as $n \to \infty$ if $\limsup_{n \to \infty} |f(n)|/g(n) < \infty$. Furthermore, $f(n) = \Omega(g(n))$ as $n \to \infty$ if $\limsup_{n \to \infty} |f(n)/g(n)| > 0$ and $f(n) = \omega(g(n))$ as $n \to \infty$ if $\limsup_{n \to \infty} |f(n)/g(n)| = \infty$. Finally, $f(n) = \bigT{g(n)}$ as $n \to \infty$ if $f(n) = \bigO{g(n)}$ and $g(n) = \Omega(f(n))$.

 %In addition, $X_n \xrightarrow[n\to\infty]{L^1} X$ denotes convergence in expectation, i.e. $\Exp{|X_n - X|} \to 0$ as $n \to \infty$, which is a stronger notion than convergence in probability.

@@

\begin{theorem}\label{thm:mainktoinfty}
Let $\alpha>\frac12, \nu>0$ be fixed and let $k_n$ be a sequence of integers
satisfying $1 \ll k_n \ll n^{1/(2\alpha+1)}$. Then, writing $G_n := G(n;\alpha,\nu)$, we have
\[
	\E[|c(k_n;G_n)-\gamma(k_n)|] = \smallO{\gamma(k_n)},
\]
as $n \to \infty$, where $\gamma(\cdot)$ is as in Theorem~\ref{thm:mainkfixed}. In particular
\[
	\lim_{n \to \infty} \Exp{\left|\frac{c(k_n; G_n)}{\gamma(k_n)}- 1 \right|} = 0.
\]
\end{theorem}

@@

Note that Theorem~\ref{thm:mainktoinfty} implies that the clustering function of the KPKVB model scales as $\gamma(k)$ as $k$ grows, whose scaling is given in the above result. In particular, this contradicts the scaling conjectured in~\cite{krioukov2010hyperbolic} for $\alpha \leq \frac34$, and confirms it for $\alpha > \frac34$.

@@

\subsection{Observations}

There are a few other observation to be made regarding our results.

\paragraph{Uniform convergence.}

Our results for the local clustering function in particular imply uniform convergence of $c(k; G_n)$ for all $2 \le k \le a_n$ where $a_n \ll n^{\frac{1}{2\alpha + 1}}$. To see this let
\[
	b_n = \arg \max_{2 \le k \le a_n} \Exp{\left|\frac{c(k_n; G_n)}{\gamma(k)} - 1\right|}.
\]
Then $b_n \le a_n \ll n^{\frac{1}{2\alpha + 1}}$ and therefore by Theorem~\ref{thm:local_clustering_hyperbolic}
\[
	\lim_{n \to \infty} \max_{2 \le k \le a_n} \Exp{\left|\frac{c(k; G_n)}{\gamma(k)} - 1\right|} 
	= \lim_{n \to \infty}\Exp{\left|\frac{c(b_n; G_n)}{\gamma(b_n)} - 1\right|} = 0.
\]

\paragraph{Near maximum scaling for $k_n$.}
Our results for the clustering function in the KPKVB model are valid for any sequence of integers $k_n \to \infty$ such that $k_n \ll n^{\frac{1}{2\alpha + 1}}$. Although one would like to have results for any sequence $k_n \le n$, it turns out that $n^{1/(2\alpha + 1)}$ is the near maximum scaling for which Theorem~\ref{thm:clustering_coefficient_hyperbolic} can be true. To see why this is the case note that by definition of the local clustering function~\eqref{eq:def_clustering_function} we have that $c(k_n; G_n) = 0$ if $N_{n}(k_n) = 0$, where $N_n(k_n)$ denote the number of vertices in a KPKVB graph $G(n; \alpha, \nu)$ with degree $k_n$. It follows by Markov's inequality that for any positive function $f : \R_+ \to \R_+$
\begin{align*}
	\Exp{\left|\frac{c_{\H,n}(k_n)}{f(k_n)} - 1\right|} 
	&\ge \Exp{\left|\frac{c_{\H,n}(k_n)}{f(k_n)} - 1\right|\ind{N_n(k_n) = 0}}\\ 
	&\ge \Prob{N_{n}(k_n) = 0} \ge 1 - \Exp{N_{\H,n}(k_n)}.
\end{align*}
We shall later establish (see Lemma~\ref{lem:diff_Nk_hyperbolic_binomial_poisson}) that $\Exp{N_n(k_n)}$ scales as $n k_n^{-(2\alpha + 1)}$. Therefore if $k_n$ is such that $k_n^{-(2\alpha + 1)} n$ tends to zero as $n \to \infty$ we have that 
\[
	\lim_{n \to \infty} \Exp{N_{n}(k_n)} = 0
\]
and hence
\[
	\lim_{n \to \infty} \Exp{\left|\frac{c(k_n; G_n)}{f(k_n)} - 1\right|} 
	\ge \lim_{n \to \infty} 1 - \Exp{N_{n}(k_n)} = 1 \ne 0,
\]
for any positive function $f$. This implies that we cannot expect a result like that of Theorem~\ref{thm:mainktoinfty} to hold as soon as $k_n \gg n^{\frac{1}{2\alpha + 1}}$.

\paragraph{Transition in scaling at $\bm{\alpha = 3/4}$.}

\PvdH{We should try and add something here. Not necessarily needed for the submission to arXiv though.}

@@

The remaining sections are devoted to the proof of Theorem~\ref{thm:mainktoinfty}, which turns out to be a lot more involved. The main reason for this is that we push the possible scaling of $k_n$ to its maximum and hence a great deal of work is needed to properly control the arising error terms and make sure these are of smaller order than $\gamma(k_n)$.

Finally, the Appendix includes some technical results on Meijer's G-function, Chernoff bounds for Poisson and Binomial random variables and the code used for simulations.

### 2019-03-13 proof_overview_finite_k.tex
finished section 9.3. Fixed the binomial Poisson coupling proof
added observation section and changed the campbell-mecke section

Imagine that we have an infinite supply of i.i.d.~points $u_1, u_2, \dots$ in the hyperbolic plane $\Haa$ chosen according to the $(\alpha, R)$-quasi uniform distribution. In the standard KPKVB random graph $G(n;\alpha,\nu)$ we take $u_1,\dots, u_n$ as our vertex set and add edges between points at hyperbolic distance at most $R = 2\log(n/\nu)$. In the {\em Poissonized} KPKVB random graph $\GPo := \GPo(n;\alpha,\nu)$, we instead take $N\isd \Po(n)$, a Poisson random variable 
with mean $n$, independent of our i.i.d.~sequence of points and let the vertex set be $u_1,\dots, u_N$ and add edges 
according to the same rule as before. Equivalently, we could say that the vertex set consists of the points of a Poisson point process with intensity function $n g$, where $g$ denotes the probability density of the $(\alpha,R)$-quasi uniform distribution. That is,
\[
	g(r,\theta) = \frac{\alpha\sinh(\alpha r)}{2\pi(\cosh(\alpha R) - 1)} \cdot 1_{\{0\leq r\leq R, -\pi<\theta\leq \pi\}}.
\]

@@

\PvdH{We might want to give some more results for Poisson point processes and Poisson random variables. Such as at the end of Section 2.1 in your recent paper on Hamiltonian cycles.}

A very useful tool for analyzing subgraph counts, and their generalizations, in the 
setting of the Poissonized random geometric graphs, and in particular the Poissonized KPKVB model and the box model is the \emph{Campbell-Mecke formula}.
We use the incarnation below, which can be found in the monograph~\cite{penrose2003random}, as Theorem 1.6.

\PvdH{I found it extremely hard to see the implications of this theorem. I therefore strongly recommend using more explicit version, for instance as in the previous version of the paper, or explicitly state the consequences of this result which will be used here.}

\begin{theorem}[\cite{penrose2003random}]\label{thm:palm}
Let $\Qcal$ be a Poisson process on $\eR^d$ with intensity function $g$, and suppose that $\lambda := \int g < \infty$.
Suppose that $h(\Ycal, \Xcal)$ is a bounded measurable function, defined on pairs $(\Ycal,\Xcal)$ with $\Ycal \subseteq \Xcal \subseteq \eR^d$
and $\Xcal$ finite, such that $h(\Ycal,\Xcal) = 0$ whenever $|\Ycal| \neq j$ (for some $j \in \eN$).
Then
\[ \Ee \sum_{\Ycal \subseteq \Qcal} h( \Ycal, \Qcal) = \frac{\lambda^j}{j!} \cdot \Ee h(\{Y_1,\dots, Y_j\}, \{Y_1,\dots, Y_j\} \cup \Qcal ), \]
\noindent
where the $Y_i$ are i.i.d.~random variables that are independent of $\Pcal$ and have common probability density function
$g/\lambda$.
\end{theorem}

>>>>

Imagine that we have an infinite supply of i.i.d.~points $u_1, u_2, \dots$ in the hyperbolic plane $\Haa$ chosen according to the $(\alpha, R)$-quasi uniform distribution. In the standard KPKVB random graph $G(n;\alpha,\nu)$ we take $u_1,\dots, u_n$ as our vertex set and add edges between points at hyperbolic distance at most $R = 2\log(n/\nu)$. In the {\em Poissonized} KPKVB random graph $\GPo := \GPo(n;\alpha,\nu)$, we instead take $N\isd \Po(n)$, a Poisson random variable 
with mean $n$, independent of our i.i.d.~sequence of points and let the vertex set be $u_1,\dots, u_N$ and add edges 
according to the same rule as before. Equivalently, we could say that the vertex set consists of the points of a Poisson point process with intensity function $n g$, where $g$ denotes the probability density of the $(\alpha,R)$-quasi uniform distribution. That is,
\begin{equation}\label{eq:def_quasi_uniform_density}
	g(r,\theta) = \frac{\alpha\sinh(\alpha r)}{2\pi(\cosh(\alpha R) - 1)} \cdot 1_{\{0\leq r\leq R, -\pi<\theta\leq \pi\}}.
\end{equation}

@@

A very useful tool for analyzing subgraph counts, and their generalizations, in the 
setting of the Poissonized random geometric graphs, and in particular the Poissonized KPKVB model and the box model is the \emph{Campbell-Mecke formula}. We use a specific incarnation, which follows from Palm theory of Poisson point processes on metric spaces, see~\cite{last2017lectures}. For this consider a Poisson point process $\Pcal$ on some metric space $\mathcal{M}$ with density $\mu$ and let $\mathcal{N}$ denote the set of all possible point configurations in $\mathcal{M}$, equipped with the sigma algebra of the process $\Pcal$. Then, for any integer $k$ and measurable function $h : \R^k \times \mathcal{N} \to \R$,
\begin{equation}\label{eq:def_campbell_mecke}
	\Exp{\sum_{p_1, \dots, p_k \in \Pcal}^{\ne} h(p_1, \dots, p_k, \Pcal)}
	= \int_{\mathcal{S}} \dots \int_{\mathcal{S}} \Exp{h(x_1,\dots, x_k, \Pcal} \dd \mu(x_1) \dots \dd \mu(x_k),
\end{equation}
where the sum is over all distinct point $p_1, \dots, p_k \in \Pcal$.

%To see this result in action consider the finite box model $\Gbox$. Then 
%\[
%	\Exp{\Nbox(k)} = \Exp{\sum_{p \in \Pcal} \ind{\Dbox(p) = k}}
%	= \int_\Rcal \Prob{\Dbox(p) = k} f(x,y) \dd x \dd y.
%\]

%
%
%We use the incarnation below, which can be found in the monograph~\cite{penrose2003random}, as Theorem 1.6.
%
%\PvdH{I found it extremely hard to see the implications of this theorem. I therefore strongly recommend using more explicit version, for instance as in the previous version of the paper, or explicitly state the consequences of this result which will be used here.}
%
%\begin{theorem}[\cite{penrose2003random}]\label{thm:palm}
%Let $\Qcal$ be a Poisson process on $\eR^d$ with intensity function $g$, and suppose that $\lambda := \int g < \infty$.
%Suppose that $h(\Ycal, \Xcal)$ is a bounded measurable function, defined on pairs $(\Ycal,\Xcal)$ with $\Ycal \subseteq \Xcal \subseteq \eR^d$
%and $\Xcal$ finite, such that $h(\Ycal,\Xcal) = 0$ whenever $|\Ycal| \neq j$ (for some $j \in \eN$).
%Then
%\[ \Ee \sum_{\Ycal \subseteq \Qcal} h( \Ycal, \Qcal) = \frac{\lambda^j}{j!} \cdot \Ee h(\{Y_1,\dots, Y_j\}, \{Y_1,\dots, Y_j\} \cup \Qcal ), \]
%\noindent
%where the $Y_i$ are i.i.d.~random variables that are independent of $\Pcal$ and have common probability density function
%$g/\lambda$.
%\end{theorem}
